Predicting Fatalities in Toronto Car Crashes: Analyzing the Impact of Time, Vehicle Type, and Other Collision Determinants

### 2024-12-04 paper/paper.qmd
final edits

---
title: "Predicting Fatalities in Toronto Car Crashes: Analyzing the Impact of Time, Vehicle Type, and Other Collision Determinants"
author: 
  - Kevin Roe
thanks: "Code and data are available at: [https://github.com/Kanghyunroe/traffic_collisions/tree/main](https://github.com/Kanghyunroe/traffic_collisions/tree/main)."
date: today
date-format: long
abstract: "This study aims to predict the likelihood of fatality in motor vehicle collisions (MVCs) in Toronto from 2014 to 2023, using data on crash time, vehicle type, and collision characteristics. By analyzing the interaction between hour of the day and vehicle types, including motorcycles and pedestrians, the model identifies key elements contributing to fatal outcomes. Results indicate that late-afternoon to evening hours, particularly between 4 PM and 11 PM, show a higher likelihood of fatalities, with motorcycles and pedestrians significantly increasing fatality risk. The model achieved a high predictive accuracy, demonstrating its potential in informing safety policies and interventions in Toronto."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our analysis identifies key predictors of vehicular accident fatalities, emphasizing heightened risks for motorcyclists and pedestrians, especially during rush hours (4PM to 5PM) and late evenings (11:00 PM). Motorcycle collisions in the evening show a fatality likelihood of around 66%. Collision type variables, such as injury and property damage collisions, significantly lower fatality risks, while police division and year have minimal impact. However, variability in estimates, particularly for motorcyclist and automobile predictors, highlights uncertainty in their precise effects, underscoring the need for targeted interventions during high-risk times and for vulnerable populations, informing policy decisions.

@@

@tbl-modelresults1 and @tbl-modelresults2 presents the estimated coefficients for the predictors in our GLM model. These coefficients fit into the GLM equation, \ref{eq:logit}, allowing us to interpret the impact of each predictor on the likelihood of fatality in a vehicular accident. Most notably, motorcyclist (17.036) and pedestrian (6.809) collisions exhibit substantial positive associations with fatality occurrence, which is in line with expectations. Moreover, motorcyclist accidents during rush hour (hour 16 and 17) have the highest coefficients at 40.4 and 26.8, respectively, and during the late evening, such as at 11:00 PM (39.365), indicating that a collision involving a motorcycle during rush hour or late evening is around 66\% likely to result in a fatality. We also find similar results with pedestrians as well. The type of collision variables all have a significant negative effect on the likelihood of death, as well.However, police division and year don't seem to have a significant impact on the likelihood of death.

@@

In @tbl-modelresults1 and @tbl-modelresults2, the coefficient of 17.036 for motorcyclists indicates that they are the most at-risk group in terms of MVC fatalities. This result strongly aligns with the widely recognized vulnerability of motorcyclists in traffic accidents. Motorcycles, by their very design, providing little protection compared to other vehicles. In the event of a crash, motorcyclists are more likely to suffer severe injuries due to the lack of structural safeguards such as airbags, seat belts, or the protective frame that cars provide. Additionally, their smaller size makes them less visible to other drivers, increasing the likelihood of collisions. This heightened risk is compounded by the fact that motorcycles are often involved in high-speed crashes, particularly on highways or in urban environments with high traffic volumes. The coefficient underscores the need for targeted interventions, such as improved rider safety training, better infrastructure (e.g., dedicated motorcycle lanes), and stricter regulations on helmet use.

@@

The analysis also highlights the significant role that environmental variables—specifically, the time of day—play in MVC fatalities. During rush hours (particularly between hours 16 and 17) and late evenings (around 11:00 PM), the likelihood of a fatality, especially in motorcycle crashes, increases substantially. The 66% fatality likelihood for motorcycle crashes during these times suggests that several environmental and behavioral elements may be at play. During rush hours, the density of traffic can lead to higher stress levels among drivers, potentially contributing to reckless driving, lane changes, or a lack of attention to motorcyclists. Additionally, the rush-hour period coincides with fatigue, both for drivers who may have been on the road for extended periods and for motorcyclists who may be navigating more complex, congested traffic conditions.

@@

- **Intercept**: 9.705
- **Hour Coefficients**:
  - `hour1` through `hour23`: Varies with each hour of the day. Positive coefficients for hours between 2 PM to 11 PM suggest higher accident probability in these hours.
  - **Key Hours**:
    - `hour2`: 1.920
    - `hour3`: 2.908
    - `hour6`: 0.966
    - `hour7`: -1.673
    - `hour8`: -0.511
    - `hour21`: 1.237
    - `hour22`: 0.631
    - `hour23`: 1.572
- **Vehicle Type Coefficients**:
  - **Motorcycle**: 17.036, indicating a high likelihood of motorcycle-related collisions.
  - **Pedestrian**: 6.832, showing the significance of pedestrian involvement in accidents.
  - **Bicycle**: 3.809, showing a moderate likelihood of bicycle-related accidents.
  - **Automobile**: 2.648, indicating that automobile-related collisions are also common.
- **Collision Type Coefficients**:
  - **Injury Collision**: -34.794, indicating a strong negative association with injury-related collisions.
  - **Fail to Remain Collision**: -29.912, indicating a strong negative association with collisions where the driver fails to remain.
  - **Property Damage Collision**: -30.004, showing a significant negative impact for property damage-related collisions.
- **Police Division Coefficients**:
  - `police_divisionD12`: 0.846, indicating division-specific effects, where division D12 has a positive association with collisions.
  - `police_divisionD13`: 0.075
  - `police_divisionD14`: -2.806
  - `police_divisionD22`: 1.797
  - `police_divisionD23`: 1.193
  - `police_divisionD31`: 0.619
  - `police_divisionD32`: 1.007
  - `police_divisionD33`: 0.055
  - `police_divisionD41`: 0.989
  - `police_divisionD42`: 1.140
  - `police_divisionD43`: 0.582
  - `police_divisionD51`: 0.915
  - `police_divisionD52`: 1.465
  - `police_divisionD53`: 1.389
  - `police_divisionD55`: 0.389
  - `police_divisionNSA`: -4.040
- **Year Coefficients**:
  - `year2015`: -0.500
  - `year2016`: 0.329
  - `year2017`: 0.412
  - `year2018`: 0.165
  - `year2019`: -0.025
  - `year2020`: 0.021
  - `year2021`: 0.079
  - `year2022`: 1.947, showing a significant increase in collisions during this year.
  - `year2023`: -1.708, showing a decrease in collisions during this year.
  - `year2024`: -0.210
- **Interaction Terms**:
  - **Hour × Motorcycle**:
    - `hour1 × motorcycle`: 31.900
    - `hour2 × motorcycle`: 33.094
    - `hour3 × motorcycle`: 33.377
    - `hour4 × motorcycle`: 33.633
    - `hour6 × motorcycle`: 51.053
    - `hour7 × motorcycle`: 52.055
    - `hour8 × motorcycle`: 55.746
    - `hour9 × motorcycle`: -66.051
    - `hour10 × motorcycle`: 27.802
    - `hour12 × motorcycle`: 51.049
    - `hour13 × motorcycle`: 18.620
    - `hour14 × motorcycle`: -12.621
    - `hour15 × motorcycle`: -11.489
    - `hour16 × motorcycle`: 32.651
    - `hour17 × motorcycle`: 20.150
    - `hour18 × motorcycle`: 12.044
    - `hour19 × motorcycle`: -11.899
    - `hour20 × motorcycle`: -5.807
    - `hour21 × motorcycle`: 21.915
    - `hour22 × motorcycle`: 9.369
    - `hour23 × motorcycle`: 34.897
  - **Hour × Pedestrian**:
    - `hour1 × pedestrian`: 15.095
    - `hour2 × pedestrian`: 29.594
    - `hour3 × pedestrian`: 20.771
    - `hour4 × pedestrian`: 19.986
    - `hour5 × pedestrian`: 23.943
    - `hour6 × pedestrian`: 9.778
    - `hour7 × pedestrian`: 1.434
    - `hour8 × pedestrian`: 0.484
    - `hour9 × pedestrian`: 1.831
    - `hour10 × pedestrian`: 1.895
    - `hour11 × pedestrian`: 15.280
    - `hour12 × pedestrian`: 9.497
    - `hour13 × pedestrian`: 1.397
    - `hour14 × pedestrian`: 12.171
    - `hour15 × pedestrian`: -0.552
    - `hour16 × pedestrian`: 0.453
    - `hour17 × pedestrian`: 0.272
    - `hour18 × pedestrian`: -5.210
    - `hour19 × pedestrian`: -0.707
    - `hour20 × pedestrian`: 7.720
    - `hour21 × pedestrian`: 12.510
    - `hour22 × pedestrian`: 1.760
    - `hour23 × pedestrian`: -2.690

@@

- **R²**: 0.992, indicating excellent fit to the data.
- **Log Likelihood**: -0.082
- **Effective Log Predictive Density (ELPD)**: -16.2, with an associated standard error of 2.2.
- **Leave-One-Out Information Criterion (LOOIC)**: 32.4, with a standard error of 4.4.
- **Watanabe-Akaike Information Criterion (WAIC)**: 21.3
- **Root Mean Squared Error (RMSE)**: 0.00, indicating near-perfect predictive accuracy.

### Training Time {#sec-appendix-modelcard-time}
- **Chain 1**:
  - Warm-up: 29.99 seconds
  - Sampling: 26.11 seconds
  - Total: 56.1 seconds
- **Chain 2**:
  - Warm-up: 19.62 seconds
  - Sampling: 17.23 seconds
  - Total: 36.84 seconds
- **Chain 3**:
  - Warm-up: 19.88 seconds
  - Sampling: 19.30 seconds
  - Total: 39.18 seconds
- **Chain 4**:
  - Warm-up: 22.40 seconds
  - Sampling: 19.17 seconds
  - Total: 41.57 seconds

>>>>

---
title: "Predicting Fatalities in Toronto Car Crashes: Analyzing the Impact of Time, Vehicle Type, and Other Collision Determinants"
author: 
  - Kevin Roe
thanks: "Code and data are available at: [https://github.com/Kanghyunroe/traffic_collisions/tree/main](https://github.com/Kanghyunroe/traffic_collisions/tree/main)."
date: today
date-format: long
abstract: "This study aims to predict the likelihood of fatality in motor vehicle collisions (MVCs) in Toronto from 2014 to 2023, using data on crash time, vehicle type, and collision characteristics. By analyzing the interaction between hour of the day and vehicle types, including motorcycles and pedestrians, the model identifies key elements contributing to fatal outcomes. Results indicate that late-afternoon to evening hours, particularly between 5 PM and 11 PM, show a higher likelihood of fatalities, with motorcycles and pedestrians significantly increasing fatality risk. The model achieved a high predictive accuracy, demonstrating its potential in informing safety policies and interventions in Toronto."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

Our analysis identifies key predictors of vehicular accident fatalities, emphasizing heightened risks for motorcyclists and pedestrians, especially during rush hour and late evenings (11:00 PM). Motorcycle collisions in the evening show a fatality likelihood of around 53%. Collision type variables, such as injury and property damage collisions, significantly lower fatality risks, while police division and year have minimal impact. However, variability in estimates, particularly for motorcyclist and automobile predictors, highlights uncertainty in their precise effects, underscoring the need for targeted interventions during high-risk times and for vulnerable populations, informing policy decisions.

@@

@tbl-modelresults1 and @tbl-modelresults2 presents the estimated coefficients for the predictors in our GLM model. These coefficients fit into the GLM equation, \ref{eq:logit}, allowing us to interpret the impact of each predictor on the likelihood of fatality in a vehicular accident. Most notably, motorcyclist (10.319) and pedestrian (6.766) collisions exhibit substantial positive associations with fatality occurrence, which is in line with expectations. Moreover, motorcyclist accidents during rush hour (5 PM) have the highest coefficients at 23.656, and during the late evening, such as at 11:00 PM (37.984), indicating that a collision involving a motorcycle during rush hour or late evening is around 53\% likely to result in a fatality. We also find similar results with pedestrians as well. The type of collision variables all have a significant negative effect on the likelihood of death, as well.However, police division and year don't seem to have a significant impact on the likelihood of death.

@@

In @tbl-modelresults1 and @tbl-modelresults2, the coefficient of 10.319 for motorcyclists indicates that they are the most at-risk group in terms of MVC fatalities. This result strongly aligns with the widely recognized vulnerability of motorcyclists in traffic accidents. Motorcycles, by their very design, providing little protection compared to other vehicles. In the event of a crash, motorcyclists are more likely to suffer severe injuries due to the lack of structural safeguards such as airbags, seat belts, or the protective frame that cars provide. Additionally, their smaller size makes them less visible to other drivers, increasing the likelihood of collisions. This heightened risk is compounded by the fact that motorcycles are often involved in high-speed crashes, particularly on highways or in urban environments with high traffic volumes. The coefficient underscores the need for targeted interventions, such as improved rider safety training, better infrastructure (e.g., dedicated motorcycle lanes), and stricter regulations on helmet use.

@@

The analysis also highlights the significant role that environmental variables—specifically, the time of day—play in MVC fatalities. During rush hour and late evenings (around 11:00 PM), the likelihood of a fatality, especially in motorcycle crashes, increases substantially. The 53% fatality likelihood for motorcycle crashes during these times suggests that several environmental and behavioral elements may be at play. During rush hours, the density of traffic can lead to higher stress levels among drivers, potentially contributing to reckless driving, lane changes, or a lack of attention to motorcyclists. Additionally, the rush-hour period coincides with fatigue, both for drivers who may have been on the road for extended periods and for motorcyclists who may be navigating more complex, congested traffic conditions.

@@

- **Intercept**: 13.146
- **Hour Coefficients**:
  - `hour1` through `hour23`: Varies with each hour of the day. Positive coefficients for hours between 2 PM to 11 PM suggest higher accident probability in these hours.
  - **Key Hours**:
    - `hour2`: 1.449
    - `hour3`: 3.851
    - `hour6`: 0.350
    - `hour7`: 0.233
    - `hour8`: 0.080
    - `hour21`: -0.184
    - `hour22`: 0.821
    - `hour23`: 2.342
- **Vehicle Type Coefficients**:
  - **Motorcycle**: 10.32, indicating a high likelihood of motorcycle-related collisions.
  - **Pedestrian**: 6.77, showing the significance of pedestrian involvement in accidents.
  - **Bicycle**: 3.45, showing a moderate likelihood of bicycle-related accidents.
  - **Automobile**: -0.26 , indicating that automobile-related collisions are also common.
- **Collision Type Coefficients**:
  - **Injury Collision**: -34.794, indicating a strong negative association with injury-related collisions.
  - **Fail to Remain Collision**: -29.912, indicating a strong negative association with collisions where the driver fails to remain.
  - **Property Damage Collision**: -30.004, showing a significant negative impact for property damage-related collisions.
- **Police Division Coefficients**:
  - `police_divisionD12`: 0.06, indicating division-specific effects, where division D12 has a positive association with collisions.
  - `police_divisionD13`: 0.17
  - `police_divisionD14`: -1.00
  - `police_divisionD22`: -0.09
  - `police_divisionD23`: 0.93
  - `police_divisionD31`: -0.65
  - `police_divisionD32`: 0.21
  - `police_divisionD33`: 0.14
  - `police_divisionD41`: 0.12
  - `police_divisionD42`: 0.46
  - `police_divisionD43`: 0.38
  - `police_divisionD51`: -0.34
  - `police_divisionD52`: 4.78
  - `police_divisionD53`: -0.88
  - `police_divisionD55`: -0.93
  - `police_divisionNSA`: -3.24
- **Year Coefficients**:
  - `year2015`: 1.03
  - `year2016`: 1.47
  - `year2017`: 1.18
  - `year2018`: 0.93
  - `year2019`: 0.55
  - `year2020`: 0.61
  - `year2021`: 1.85
  - `year2022`: 3.17, showing a significant increase in collisions during this year.
  - `year2023`: 0.14
  - `year2024`: 0.75
- **Interaction Terms**:
  - **Hour × Motorcycle**:
    - `hour1 × motorcycle`: 33.237
    - `hour2 × motorcycle`: 35.448
    - `hour3 × motorcycle`: 35.467
    - `hour4 × motorcycle`: 36.943
    - `hour6 × motorcycle`: 57.233
    - `hour7 × motorcycle`: -6.889
    - `hour8 × motorcycle`: 9.962
    - `hour10 × motorcycle`: 29.521
    - `hour12 × motorcycle`: 54.822
    - `hour13 × motorcycle`: 19.375
    - `hour14 × motorcycle`: 24.927
    - `hour15 × motorcycle`: 23.135
    - `hour16 × motorcycle`: -5.637
    - `hour17 × motorcycle`: 23.656
    - `hour18 × motorcycle`: -5.293
    - `hour19 × motorcycle`: 24.156
    - `hour20 × motorcycle`: -2.714
    - `hour21 × motorcycle`: -3.445
    - `hour22 × motorcycle`: 11.598
    - `hour23 × motorcycle`: 37.984
  - **Hour × Pedestrian**:
    - `hour1 × pedestrian`: 15.758
    - `hour2 × pedestrian`: 29.368
    - `hour3 × pedestrian`: 18.546
    - `hour4 × pedestrian`: 20.723
    - `hour5 × pedestrian`: 22.835
    - `hour6 × pedestrian`: 1.713
    - `hour7 × pedestrian`: 17.229
    - `hour8 × pedestrian`: -2.876
    - `hour9 × pedestrian`: 0.983
    - `hour10 × pedestrian`: 0.973
    - `hour11 × pedestrian`: 9.320
    - `hour12 × pedestrian`: -2.071
    - `hour13 × pedestrian`: -1.027
    - `hour14 × pedestrian`: 12.206
    - `hour15 × pedestrian`: -0.145
    - `hour16 × pedestrian`: 0.042
    - `hour17 × pedestrian`: -0.494
    - `hour18 × pedestrian`: -0.977
    - `hour19 × pedestrian`: 0.090
    - `hour20 × pedestrian`: 1.757
    - `hour21 × pedestrian`: -0.436
    - `hour22 × pedestrian`: 14.945
    - `hour23 × pedestrian`: 12.795

@@

- **R²**: 0.992, indicating excellent fit to the data.
- **Log Likelihood**: -0.326
- **Effective Log Predictive Density (ELPD)**: -17.6, with an associated standard error of 3.1.
- **Leave-One-Out Information Criterion (LOOIC)**: 35.1, with a standard error of 6.2.
- **Watanabe-Akaike Information Criterion (WAIC)**: 21.3
- **Root Mean Squared Error (RMSE)**: 0.00, indicating near-perfect predictive accuracy.

### Training Time {#sec-appendix-modelcard-time}
- **Chain 1**:
  - Warm-up: 24.63 seconds
  - Sampling: 18.63 seconds
  - Total: 43.26 seconds
- **Chain 2**:
  - Warm-up: 22.78 seconds
  - Sampling: 26.92 seconds
  - Total: 49.70 seconds
- **Chain 3**:
  - Warm-up: 31.42 seconds
  - Sampling: 24.05 seconds
  - Total: 55.47 seconds
- **Chain 4**:
  - Warm-up: 26.75 seconds
  - Sampling: 27.47 seconds
  - Total: 54.21 seconds
